\documentclass[a4paper,12pt]{report}

\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{xurl}
\usepackage{array}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{setspace}
\usetikzlibrary{shapes.geometric, arrows}

\doublespacing

\title{How Has the Development of Digital Cryptography Both Influenced and Been
Influenced by Moral and Legal Discussions Surrounding the Right to Privacy?}
\author{Alex Ydens}

% LATEX ARTICLES:
% - https://latex-tutorial.com/tutorials/
% - https://www.overleaf.com/learn/latex/
% - https://www.overleaf.com/learn/latex/LaTeX_Graphics_using_TikZ%3A_A_Tutorial_for_Beginners_(Part_1)%E2%80%94Basic_Drawing
% LATEX NOTES:
% - \section{Section Name} for a section header
% - \subsection{SubSection Name} for a subsection header
% - \paragraph{Paragraph Caption} for a captioned paragraph
% - \subparagraph{Subparagraph Caption} for a captioned subparagraph
% -  surround inline equations in '$' symbols
% - \begin{equation} \end{equation} for multiline equations
% - \begin{equation*} \end{equation*} for non numbered multiline equations
% - \begin{align} \end{align} for multiline equations aligned at the '&' symbols
% - \begin{align*} \end{align*} for non numbered multiline equations aligned at
% the '&' symbols
% -  '\\' in a multiline equation for a new line
% - \label{ident} to label something, \ref{ident} to reference it
% - \left[ \right] for scaled '[]' brackets. same works for '(){}'
% - \begin{matrix} \end{matrix} (with '\\' and '&') for matrices
% - \begin{figure} \end{figure} for a figure
% - \begin{figure}[vals]: '!' for force, 'h' here, 't' top, 'b' bottom, 'p' own
% page. This works for table too, and probably lost of other things.
% - \begin{subfigure}[a-z] \end{subfigure}[a-z] for a subfigure with a label
% (\begin{subfigure}[a] will make figure N(a))
% - \caption{Caption} for a figure caption
% - \label{fig:ident} for a figure label
% - \cite{ident} for a citation
% - \footnote{Footnote} for a footnote
% - \begin{table} \end{table} for a table
% - \url{https://xkcd.com} for a url
% - \textbf{ABC} for bold text
% - \textit{ABC} for italics text
% - \begin{itemize} \end{itemize} with lots of \item for bullet list
% - \begin{enumerate} \end{enumerate} with lots of \item for bullet list

%   \begin{minted}{c}
% #include <stdio.h>
% 
% int main(int argc, char *argv[]) {
%   printf("Hello world!\n");
%   return 0;
% }
% 
%   \end{minted}
% \inputminted{c}{../src/main.c}
% \mintinline{c}|printf("Hello %d\n", 1);|

% EPQ NOTE: 5,000 words target (can be slightly over)

\begin{document}
  \frenchspacing
  \pagenumbering{gobble}
  \maketitle
  \newpage
  \tableofcontents

  \newpage
  \pagenumbering{arabic}
  \section{Abstract}

  \newpage
  \section{Introduction}
These days, trust and security over the internet is becoming more and more
essential. From communicating with banks, to managing sensitive information in a
business, to communicating privately over a service such as WhatsApp, two things
are always necessary: being able to verfiy you are who you say you are, and
being able to send a message that cannot be intercepted and understood in
transit. To accomplish this goal, digital cryptography has had to make several
crucial advances over the last 70 years, but it has been far from smooth
sailing. With so much valuable information at stake, and the risks of giving
people absolute secrecy, research into digital cryptography has marked by
frequently changing attitudes, standards and restrictions. The effect of these
external factors on this field of study, and the implications some of the
advances in it have yielded will be the focus of this document.

  \section{Symmetric Encryption}
The first major breakthrough in digital encryption actually came before
computers in their current form existed, in the 1910s, in the form of the Vernam
cipher. The Vernam cipher, patented by Gilbert Vernam in 1919\cite{vernam}, was
one of the earliest examples of a truly unbreakable cipher, when used correctly.

    \subsection{Using the Vernam Cipher}
To send a message using the Vernam cipher, the sender must first ensure their
message uses a binary encoding. On computers, this is trivial - all data is
encoded in binary on a digital computer - but when originally created, this
required using a tape with holes in it to indicate zeros. Then, a key, which
consists of a binary string of a length greater than or equal to that of the
message, must be bitwise-XORed with the message to obtain the ciphertext. On the
teletype machines of Vernam's time, this required laying the two tapes on top of
each other, and reading any position which had a hole through one tape but not
the other as a one, and all other positions as a zero. To get the plaintext from
the ciphertext, the process could be done in again - just perform a
bitwise-XOR on the ciphertext with the key, and the plaintext should emerge. See
the table below for proof this process works as intended.
      \begin{center}
      \begin{table}[h!]
      \begin{tabular}{ |c|c|c|c| }
        \hline
Plaintext & Key & Ciphertext (Plaintext XOR Key) & Plaintext (Ciphertext XOR
Key) \\
        \hline
0 & 0 & 0 & 0 \\
        \hline
0 & 1 & 1 & 0 \\
        \hline
1 & 0 & 1 & 1 \\
        \hline
1 & 1 & 0 & 1 \\
        \hline
      \end{tabular}
      \caption{Verifying the Vernam Cipher works as intended.}
      \end{table}
      \end{center}
    \subsection{Issues with the Vernam Cipher}
Despite being perfect in theory, and working well with the binary encoding used
by digital computers, the Vernam Cipher has some fairly serious practical issues
when used in isolation. First of all, to ensure perfect secrecy, the key must be
at least as long as the message, completely random, and only known to the sender
and the reciever. This means that, for each message sent, a sender and reciever
would first have to generate a key as long as the message and send it privately.
If the option of privately sending a bit pattern as long as the message was
already available, then the sender and the reciever would not have needed to use
encryption in the first place, and as such the Vernam cipher seems to have
delegated the problem of sending secret messages to key distribution, rather
than having solved the issue.

That being said, initially, there was simply no better way when sending private
digital messages, meaning that those who could afford the cost and were able to
handle the logistics (primalriy large organisations), hired people to deliver
keys in groups of several at a time to those they wanted to communicate with,
and then used the cipher with the delivered keys\cite{singh}. It wasn't perfect,
as the keys could be lost in transit and more would be needed every so often,
but it was better than sending messages regarding sensitive business matters in
plaintext, where anyone without much effort could intercept and read them.

  \section{The DES}
    \subsection{Development}
In response to the increasing use of digital encryption, and a desire to have a
standardised way to utilise it in the government, America's National Bureau of
Standards began to work in collaboration with the public to form a standard for
digital cryptography that could be implemented by whoever needed it, simplifying
the process of cross-organisational communication over encrypted channels.

This was formed initially of an algorithm built on similar principles as
Vernam's cipher created at IBM by Horst Feistel, built into a cipher system that
would hold up well against the attacks of the time. This wasn't that much more
advanced as the Vernam and held the same issues as it, but that could be said of
anything on the market at the time.

The standard was then ran by the National Security Agency of America, who made
some changes. Firstly, they modified it to be less vunerable to differential
cryptanalysis, but more vunerable to brute-force methods, with a decreased key
length of 56 bits\cite{schneier}. This decision was motiviated to ensure that
DES could not be broken by most people, but organisations would be able to put a
lot of time and effort into brute forcing it, thus allowing the US government
the ability to break their citizen's encryption as they saw fit, and ensuring
the standard they released could not be used against them be foreign
adversaries. After this, in 1977, it was released.
    \subsection{Criticism}
The decisision to lower the key length did not go unnoticed amongst the US
public, however. For example, Martin Hellman, who would go on to be one of the
leading figures behind asymmetric encryption, initially spotted that the DES was
vunerable to brute force attacks and, not knowing that this was a conscious
decision by his government, raised the issue to the people behind the standards,
only to be met with a response explaining the reasoning behind the decision. He
made his dissapointment about this clear in a later lecture, stating ``What we
had thought was a technical problem turned out to be political. If we wanted to
improve the security of the standard, we would have to treat it as a political
battle by seeking media coverage and congressional hearings - which we did.".

And Hellman et al were not alone in this. Many critics did not see the weakening
of the security of their communication as a move that would help protect the US
public, seeing as foreign threats could easily develop their own standards.
Furthermore, critics argued that the government having the power to read their
secrets encroached on their right to privacy, treating people as guilty until
proven innocent and allowing for mass surveillance of communications.

It was this frustration with the US government's position on encryption,
according to some\cite{schneier}, that lead to much of the innovation that was
to come in the following years. It certainly was an inspiration for Martin
Hellman\cite{hellman}.

  \section{Diffie-Hellman-Merkle Key Exchange and GCHQ's Equivalent}
The issue of key distribution for long keys, and the insecurity of shorter-key
algorithms such as the one used for the DES standard gave rise to many people to
work on solutions to the key exchange problem, where each solution had its fair
share of upsides and downsides.
    \subsection{Merkle's Puzzles}
Quite possibly the first viable solution to this problem was devised by Ralph
Merkle in 1974\cite{merkle}. The idea was simple: have the recipient share a
piece of information that can't be used to decipher a message but can be used to
encipher one, so that a sender can encipher their message and send it without
requiring any previous secret communcation, and the reciever can then use the
information they used to generate the public information to decipher the
message. In this way, the key distribution problem is far easier to solve, as
the information containing the keys can be sent publicly without fear of
interception.

The idea was to follow the steps outlined in Figure \ref{fig:merklepuzzle}.

    \begin{figure}[h!]
      \begin{enumerate}
\item First, the reciever encrypts \(2^N\) messages of the form: ``The message
is X, the key is Y" with an algorithm that can be brute forced by a sender with
some difficulty, such that brute forcing one would be practical, but brute
forcing \(2^N\) would not. For each message, \(X\) and \(Y\) are chosen
randomly, and the key used to encrypt the message is \(Y\). These messages are
shared publicly, for anyone to use to send their message.
\item The sender then takes a random message from those the recierver has sent,
and uses a brute force method to decrypt it. When they have achiever this, they
send back their message encrypted by the key \(Y\) from the message they
decrypted, followed by the value for \(X\) from that message in plaintext.
\item The reciever can now take the message sent, look through the plaintext
forms of the messages they had encrypted and sent, and find the one which
references the same plaintext \(X\) at the end of the sender's message. This
will point them to the correct \(Y\) to decrypt and read the message.
\item Without knowing which of the \(2^N\) messages the sender chose to brute
force to send their message, an interceptor would need to keep brute forcing
them until the correct one is found, which is impractical.
      \end{enumerate}
      \caption{Steps required to use Merkle's Puzzles.}
      \label{fig:merklepuzzle}
    \end{figure}

With this system in place, a sender can send the information necessary for
anyone to encrypt a message and send it to them freely without worrying that an
interceptor of this will be able to decipher the following message. As such, key
distribution does not have to take place in secret anymore, so keys can be
bigger, and the Vernam cipher can be used, strengthening messages considerably.

There are two main issues with this system. First of all, it is rather
impractical. To ensure brute forcing \(2^N\) messages cannot occur in a
reasonable time frame, the sender needs to ensure that brute forcing one takes a
considerable amount of time, making the process required of the sender
computationally taxing. Secondly, it is possible the interceptor could get
lucky, and the message the sender chose is one of the first the interceptor
chose to brute force, causing the message to be read without much hassle. As
such, Merkle's Puzzles gave the field of cryptography research hope that the key
distribution problem was solvable, but did not yet provide a solution that would
see much use.

    \subsection{Diffie-Hellman(-Merkle) Key Distribution}
A subsequent approach to this, based on a similar principle, is still in use
today for certain applications. It is most commonly referred to as the
Diffie-Hellman key disitrbution algorithm, although in this document I will call
it the Diffie-Hellman-Merkle key distribution algorithm, which is what Martin
Hellman preffers it be called, in an effort to acknowledge the Ralph Merkle's
earlier contributions\cite{hellman}. The exact method requires some tricky
number theory and modular arithmetic\cite{diffiehellman}, but the idea can be
shown using the analogy given in Figure \ref{fig:diffiehellman}.
    \begin{figure}[h!]
      \begin{enumerate}
\item Let's say person A and person B both have a common starting paint,
available to anyone, and want to end up with a common ending paint that no one
but them can recreate perfectly.
\item Person A and person B can both select a secret colour, and mix it into
their paint. They can then send this mixture publicly, knowing that paint can't
be unmixed, so neither person's secret colour can be inferred from the mixture.
This allows them to safely swap their paint mixtures.
\item Next, they take the mixture they recieved and add their secret colour to
it. As such, each mixture now contains the starting paint, person A's secret
colour, and person B's secret colour - both people have the same secret paint
that nobody can replicate, without having to have send it securely to one
another.
      \end{enumerate}
      \caption{Steps required to use Merkle's Puzzles.}
      \label{fig:diffiehellman}
    \end{figure}
Because this idea only required modular arithmetic and raising integers to
integer powers, it was easy and quick to implement on a computer, and could
provide unparalelled security. As long as person A and person B carried out
their numerical equivalent to the process outlined in Figure
\ref{fig:diffiehellman}, they could be sure that they both had a key no one else
could have guessed from the information they exchanged, and could proceed with
symmetric encryption to transfer their message.

That being said, this approach is not completely without its downsides. Mainly,
it requires both the sender and the receiver to actively exchange information
and carry out computations before sending a message, making the process of
sending information somewhat tedious. Ideally, a reciever would only have to
share information once, and a sender could send as many messages as they want to
them without issue.

    \subsection{GCHQ's Equivalent Discovery}
Diffie-Hellman-Merkle key exchange had actually been discovered before it was
independently discovered by Diffie and Hellman and shared with the public, in
GCHQ. Clifford Cocks and James H. Ellis had collaborated on the project after
being shown the problem and becoming interested in it in the mid 1970s, but were
not able to release their findings to the public because it was kept secret by
GCHQ until December 1997\cite{ellis}.

This prevented this the original discoverers from getting the credit they
deserved in their lifetimes, and lead to the advancements in the field being
delayed by some time. Furthermore, GCHQ is not known to have used these findings
themselves, not realising how revolutionary they were at the time. As such, the
government's secrecy surrounding this discovery impeded research and did not
benefit anyone.

  \section{RSA Encryption}
A system that solves this issue, and has been the encryption method of choice on
the internet for several decades, is known as the RSA cryptosystem, after its
creators, Ron Rivest, Adi Shamir and Leonard Adleman\cite{rivestshamiradleman}.
Despite people moving away from it as of recent, the aproach it uses has become
the foundation for most of the algorithms that have followed it, and is known as
public-key cryptography.

The idea is as follows. If a cipher could be devised such that a different key
is required to encrypt it than the key required to decrypt it, and it is
suitably difficult to generate the decryption key from the encryption key, then
a receiver can just send out their encryption key and let anyone encrypt a
message with it, and keep their decryption key private and use it to decrypt
messages. In this way, once encrypted, not even the sender can read the message
- only the receiver, that has the decryption key, can do that. Usually, the key
used for decryption is known as the private key, as it should not be shared lest
the system be compromised, and the one used for encryption is known as the
public key, as it can be shared with anyone who wishes to send a message to the
reciever. This is the principle of public-key cryptography, and a cipher that
has a different key for encryption than the one it uses for decryption is known
as an asymmetric cipher.

The advantages of this system are massive over that of the Diffie-Hellman-Merkle
key exchange are clear to see. When using the Diffie-Hellman-Merkle key
exchange, any time a sender wants to send a message, it must coordinate a key
exchange process with the sender, with both sides needing to send various
numbers and perform various calculations, and this process having to be repeated
for each new message. When it comes to using a public key, however, the process
is far simpler. The reciever has to perform some calculations once, and can then
send out a key that can be used for all future communications, for all future
senders. The receiver need only recieve their key - no calculation is required
on their end.

In the case of the RSA cryptosystem in particular, the exact method is out of
the scope of this document, but it is relies on the fact that if a private key
is comprised of two sufficiently large primes, and a public key is comprised of
their product, finding the prime factors of that product, and thus determining
the private key from the public key, is not computationally feasable on a modern
computer.

  \section{Restrictions on RSA and PGP's Legal Issues}
Initially, the US government were not pleased with the discovery of public-key
encryption due to the fact that it made it significantly harder for them to
wire-tap individuals they deemed suspicious. As such, just as the DES had a
reduced key length to make it more prone to brute force attacks, RSA with a
large enough key was considered a munitions, and so could not be distributed
without prosecution for arms dealing. This revitalised national debate in the US
over the right to privacy, with many protesting the US government's strict
stance on the matter, arguing it wasn't their business what they chose to send
to each other, and that criminals had no reason to comply to the regulations, so
the main victims would be law-abiding citizens for whom wire tapping is wholly
unnecessary. It took the US government until 1996 to take a more relaxed stance
on the matter, removing encryption from their list of munitions, considering it
a commercial product instead\cite{clinton}.

    \subsection{PGP}
When RSA encryption was first created, it was patented by its creators, who
formed a company to provide encryption to those who required it. This company is
still in business today, under the name RSA Security USA LLC\cite{rsa}. This
meant that, intially, the algorithm was not to be used or sold without acquiring
a license from the company. However, Phil Zimmermann, an American computer
scientist at the time, created his own implementation of the cryptosystem,
wrapped up in an easy to use application, and titled it ``Pretty Good Privacy",
or PGP for short. After sending it to a friend, it quickly spread across Usenet
forums, allowing anyone to use the algorithm without a license\cite{zimmermann}.

RSA Security USA LLC were understandably not pleased. If potential customers
could get an implementation of the same product they were offering, wrapped up
in an easy to use package and already in wide use completely for free, they were
exceedingly unlikely to require an implementation that came with a recurring
cost and was used by (and therefore supported by) comparitively less people. As
such, in May 1997, RSA Security USA LLC took Phil Zimmermann to
court\cite{leaks}. This case was resolved through a settlement.

This was not the end of PGP's legal troubles, however. Due to the fact strong
encryption was considered a munition by the US government at the time of PGP's
release, Zimmermann was investigated by the US government for arms
dealing\cite{dubois}. The US government's argument was that since the source
code had been distributed to several countries without consulting the
government, and it was considered a munition, this constituted arms dealing of
Zimmermann's part. Zimmermann responded in an unorthodox way, publishing the
full source code of the software in a hardback book. The reasoning behind this
was that the while the export of munitions was controlled by the US government,
the export of books was protected by the First Ammendment, and so controlling
the export of the book was much harder to justify legally. Furthermore, by
publishing it as a book, Zimmermann attempted to frame the classification as
ridiculous, as a book cannot be used as a weapon, and was being treated the same
as bombs and guns. In the end, through a mixture of the changing regulations,
the US public being firmly in Zimmermann's favour, and the fact that it was
difficult to legally restrict a book, the investiagion was closed without any
charges being filed.

  \section{Quantum Computing's Impact on Digital Cryptography}

  \section{Standards for Post-Quantum Cryptography}

  \newpage
  \section{Appendix}
    \begin{appendix}
      \listoffigures
      \listoftables
      \bibliography{epq}
      \nocite{*}
      \bibliographystyle{ieeetr}
    \end{appendix}
\end{document}
